---
title: "Story6_Data608"
format: revealjs
editor: visual
---

## Defining Business Question

This assignment involves using dimensionality reduction, clustering, and other tools to tell a story about the types of customers at Instacart.

```{r, warning = FALSE, message = FALSE, echo=FALSE, include=FALSE}
# echo=FALSE, include=FALSE

# Load required libraries

library(tidyverse)
library(caret)
library(ggplot2)
library(corrplot)
library(GGally)
library(randomForest)
library(gridExtra)
library(pscl)
library(pROC)
library(MASS)
library(boot)
library(Metrics)
library(dplyr)
library(Rtsne)
library(cluster)
library(tibble)     # for tibble conversions
library(tidyr)      # for reshaping data (preferred over reshape2)
library(forcats)    # for better category handling in plots
library(tibble)

# Loading Dataset
aisles_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_12/aisles.csv")
all_order_products_df <- read_csv("C:/Users/Uzma/CUNY-SPS-Assignments/Data_608/Week_12/all_order_products.csv")
departments_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_12/departments.csv")
orders_df <- read_csv("C:/Users/Uzma/CUNY-SPS-Assignments/Data_608/Week_12/departments.csv")
products_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_12/products.csv")
user_features_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_12/user_features.csv")
head(user_features_df)

```

## Reducing Customer Dimensions to Reveal Behavior Patterns

```{r fig.width=10, fig.height=6}

# Identify feature types
all_cols <- colnames(user_features_df)
day_cols <- all_cols[str_detect(all_cols, "^(Saturday|Sunday|Monday|Tuesday|Wednesday|Thursday|Friday)$")]
product_cols <- setdiff(all_cols, c("user_id", day_cols))

features <- user_features_df %>% dplyr::select(all_of(product_cols))
features_scaled <- scale(features)

pca_result <- prcomp(features_scaled, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2]
)

# Plot with annotation
ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.6, color = "#2c7fb8") +

  # Add annotation box
  annotate("text", x = min(pca_df$PC1) + 10, y = max(pca_df$PC2) - 10,
           label = "PCA reveals customer similarity:\nClose points = similar shopping habits",
           size = 4, fontface = "italic", color = "black", hjust = 0) +

  labs(
    title = "Customer Behavior Patterns via PCA",
    subtitle = "Each point represents a customer’s shopping behavior",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )


```

## Clustering Users Based on Shopping Behavior

```{r fig.width=10, fig.height=6}
user_features_df <- as_tibble(user_features_df)
# select product columns
features <- user_features_df %>%
  dplyr::select(-user_id, -Saturday, -Sunday, -Monday, -Tuesday, -Wednesday, -Thursday, -Friday)

features_scaled <- scale(features)  # <- THIS is what we use in kmeans


pca_result <- prcomp(features_scaled, center = TRUE, scale. = TRUE)
pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2]
)

set.seed(42)
k_result <- kmeans(features_scaled, centers = 4)

# Add cluster labels
pca_df$cluster <- k_result$cluster

# Replace numeric cluster IDs with meaningful labels
cluster_labels <- c(
  "1" = "Cluster 1: Mixed Shoppers",
  "2" = "Cluster 2: Health-Focused",
  "3" = "Cluster 3: Snack/Convenience",
  "4" = "Cluster 4: Niche/Low Activity"
)

# Add descriptive labels to a new column
pca_df$cluster_label <- factor(cluster_labels[as.character(pca_df$cluster)],
                               levels = cluster_labels)  # keeps legend order

# Plot with legend
ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster_label)) +
  geom_point(alpha = 0.8, size = 1.8) +
  labs(
    title = "Customer Segments Identified by Clustering",
    subtitle = "K-Means identified 4 major customer types",
    x = "PC1", y = "PC2",
    color = "Customer Cluster"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  scale_color_manual(values = c(
    "Cluster 1: Mixed Shoppers" = "#e41a1c",
    "Cluster 2: Health-Focused" = "#4daf4a",
    "Cluster 3: Snack/Convenience" = "#377eb8",
    "Cluster 4: Niche/Low Activity" = "#984ea3"
  ))



```
## What Each Cluster Buys (Top Product Categories)

```{r}
# Add cluster back to user features
clustered_users <- user_features_df %>%
  mutate(cluster = k_result$cluster) %>%
  dplyr::select(cluster, all_of(product_cols))

# Convert to long format
long_cluster_data <- clustered_users %>%
  pivot_longer(cols = -cluster, names_to = "category", values_to = "count") %>%
  group_by(cluster, category) %>%
  summarise(avg = mean(count), .groups = "drop")

# Top 10 categories overall
top_cats <- long_cluster_data %>%
  group_by(category) %>%
  summarise(total_avg = sum(avg)) %>%
  slice_max(total_avg, n = 10) %>%
  pull(category)

# Filter to top categories
filtered_plot_data <- long_cluster_data %>%
  filter(category %in% top_cats)

# Plot
ggplot(filtered_plot_data, aes(x = avg, y = fct_reorder(category, avg), fill = factor(cluster))) +
  geom_col(position = "dodge") +
  labs(
    title = "Top Product Categories by Cluster",
    subtitle = "Average items purchased per user, across top 10 categories",
    x = "Avg Items", y = "Product Category", fill = "Cluster"
  ) +
  theme_minimal(base_size = 13) +
  scale_fill_manual(values = c("#e41a1c", "#4daf4a", "#377eb8", "#984ea3"))

```
## Shopping Behavior Over Weekdays (Dynamic Day Columns)

```{r}
# Recode cluster labels for legend clarity
cluster_labels <- c(
  "1" = "Cluster 1: Mixed Shoppers",
  "2" = "Cluster 2: Health-Focused (Midweek)",
  "3" = "Cluster 3: Weekend Snackers",
  "4" = "Cluster 4: Low-Activity Users"
)

# Prepare long-format data with dynamic days
day_cols <- grep("^(Monday|Tuesday|Wednesday|Thursday|Friday|Saturday|Sunday)$",
                 colnames(user_features_df), value = TRUE)

long_days <- user_features_df %>%
  mutate(cluster = k_result$cluster) %>%
  dplyr::select(cluster, all_of(day_cols)) %>%
  pivot_longer(cols = all_of(day_cols), names_to = "day", values_to = "count") %>%
  group_by(cluster, day) %>%
  summarise(avg_items = mean(count), .groups = "drop") %>%
  mutate(
    cluster = factor(cluster_labels[as.character(cluster)],
                     levels = cluster_labels),
    day = factor(day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
  )

# Plot with annotated legend
ggplot(long_days, aes(x = day, y = avg_items, color = cluster, group = cluster)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Weekly Shopping Pattern by Cluster",
    subtitle = "Shopping habits vary in timing across customer types",
    x = "Day of Week", y = "Avg Items Purchased",
    color = "Customer Type"
  ) +
  scale_color_manual(values = c(
    "Cluster 1: Mixed Shoppers" = "#e41a1c",
    "Cluster 2: Health-Focused (Midweek)" = "#4daf4a",
    "Cluster 3: Weekend Snackers" = "#377eb8",
    "Cluster 4: Low-Activity Users" = "#984ea3"
  )) +
  theme_minimal(base_size = 13) +
  theme(
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

