---
title: "Story3_Data608"
author: "Mubashira Qari"
date: 2025-2-23"
format: revealjs
editor: visual
---

```{r, warning = FALSE, message = FALSE, }
# echo=FALSE, include=FALSE

library(dplyr)
library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(ggfortify)
library(RColorBrewer)
library(tidyr)
library(ggplot2)
library(dplyr)
library(usmap)   # For U.S. map plotting
library(sf)      # For spatial data handling 
library(viridis) # For color scales


# Loading Dataset

# Code book: https://brightspace.cuny.edu/content/enforced/680560-SPS01_DATA_608_1252_18469/37363-0001-Codebook-ICPSR.pdf

state_overview_df <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/State_Overview-2019-2023.csv")

firearm_laws_df <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/firearm_laws.csv")

mortality_breakdown_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/mortality_breakdown.csv")

firearm_mortality_data <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/firearm_mortality2022.csv")

state_abbreviation_df <- read_csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/states_abbreviations.csv")

#mortality_breakdown_df
#state_abbreviation_df
#state_overview_df
#firearm_mortality_data
#firearm_laws_df

```



```{r}
### Preparing for API Call Census 2010 Data
Sys.setenv(census_api_key = "f7ce5a76ddf2088c73fda9c0a87410995cebbffa")
#Sys.getenv("census_api_key")
```


```{r}
# Access the API key
# Save the Census API key in the environmental variable.
#census_api_key("YOUR KEY GOES HERE", install = TRUE)
```


```{r}
#Sys.getenv("census_api_key")
```


```{r, warning = FALSE}
#, message = FALSE, echo=FALSE, include=FALSE
# Fetching Census Data for the year 2020
# https://api.census.gov/data/2020/acs/acs1?get=NAME,B01001_001E&for=state:*
# Load required packages

library(tidycensus)
library(dplyr)

census_df <- get_acs(
  geography = "state",
  variables = c(population = "B01003_001E",
                median_age = "B01002_001E",
                household_income = "B19013_001E",
                per_capita_income = "B19301_001E",
                poverty_count = "B17001_002E",
                unemployment_count = "B23025_005E"
                ),
  output = "wide",
  year = 2020
)

#head(census_df)
# View the actual column names in census_df
#colnames(census_df)

```

```{r}
# Rename column name
firearm_mortality_data <- firearm_mortality_data %>%
  rename(CODE = STATE)

# Replace a value in a column

firearm_mortality_data <- firearm_mortality_data %>%
  mutate(CODE = ifelse(CODE == "District of Columbia", "DC", CODE))


library(dplyr)
# Rename column name 
mortality_breakdown_df <- mortality_breakdown_df %>%
  rename(code = State)


print(mortality_breakdown_df)

# Rename all column names to lowercase
library(dplyr)

firearm_mortality_data <- firearm_mortality_data %>%
  rename_with(tolower)

state_abbreviation_df <- state_abbreviation_df %>%
  rename_with(tolower)

state_overview_df <- state_overview_df %>%
  rename_with(tolower)

mortality_breakdown_df <- mortality_breakdown_df %>%
  rename_with(tolower)

firearm_laws_df <- firearm_laws_df %>%
  rename_with(tolower)

firearm_laws_df
```


```{r}
library(dplyr)

# Your vector of required columns
required_columns <- c("NAME", "population", "median_age", "household_income", 
                      "per_capita_income", "poverty_count", "unemployment_count")

# Now use the pipe (%>%) properly

census_df <- census_df %>%
  dplyr::select(all_of(required_columns))


# Rename all column names to lowercase

census_df <- census_df %>%
  rename_with(tolower)


print(census_df)

```

```{r}
#detach("package:MASS", unload = TRUE)
library(dplyr, warn.conflicts = FALSE)

```


```{r}

# Keeps all rows from firearm_mortality_data and fills in NA for missing matches in state_abbreviation
state_overview_df <- state_overview_df %>%
  left_join(state_abbreviation_df, by = c("state" = "state"))

mortality_breakdown_df <- mortality_breakdown_df %>%
  left_join(state_abbreviation_df, by = c("code" = "code"))

firearm_mortality_data <- firearm_mortality_data %>%
  left_join(state_abbreviation_df, by = c("code" = "code"))

firearm_laws_df <- firearm_laws_df %>%
  left_join(state_abbreviation_df, by = c("state" = "state"))

#mortality_breakdown_df
#firearm_mortality_data
#state_abbreviation_df
#state_overview_df
firearm_laws_df

```



```{r}
library(tidyverse)

#firearm_mortality_data
state_overview_df
#firearm_laws_df

```

## More Gun Laws Lead to Fewer Firearm Deaths?

-   This scatter plot will shows that states with overall more gun laws have lower firearm death rates.
-   The red regression line trends downward, it suggests that more gun laws lead to fewer firearm deaths.



```{r}
firearm_mortality_data <- firearm_mortality_data %>%
  left_join(firearm_laws_df, by = c("state" = "state"))
firearm_mortality_data

```

## Impact of Stricter Firearm Permit Laws

-   This graph shows the firearm mortality rate by state in the United States for 2022, with states ordered from highest to lowest mortality rates.


```{r}
# Filter for year 2022
firearm_data_2022 <- firearm_mortality_data %>% 
  filter(year.x == 2022)

# Bar plot of states by rate
library(ggplot2)
library(dplyr)

ggplot(firearm_data_2022, aes(x = reorder(state, -rate), y = rate, fill = permitlaw)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Stricter Firearm Permit Laws Correlate with Lower Mortality Rates Across States",
       x = "State",
       y = "Mortality Rate",
       fill = "Permit Law Status") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
## Key observations:

Mississippi has the highest firearm mortality rate (approximately 30 deaths per 100,000 population), followed by Louisiana and Alabama.
States are color-coded by their permit law status, indicated by a gradient from dark blue (0.00) to light blue (1.00). This appears to represent the strictness of gun permit laws, with lighter colors indicating stricter permit requirements.
There's a noticeable pattern where many states with higher mortality rates (mostly dark blue) have less restrictive permit laws (0.00), while states with lower mortality rates (toward the right side of the graph) tend to have more restrictive permit laws (lighter blue).
The lowest firearm mortality rates are seen in Rhode Island, Massachusetts, Hawaii, New Jersey, and New York - all of which appear to have stricter permit laws as indicated by their lighter blue coloring.
There are some exceptions to this pattern, suggesting that other factors beyond permit laws also influence firearm mortality rates.
This visualization suggests a potential correlation between stricter permit laws and lower firearm mortality rates, though correlation doesn't necessarily imply causation.



## The Regression Model Interpretation:

-   The regression model and scatter plots confirm that states with stricter gun laws (higher LAWTOTAL scores) tend to have lower firearm death rates.
-   States requiring handgun permits (PERMITH \~= 1) show fewer deaths, while states without permit requirements see higher firearm mortality.

```{r}

firearm_mortality_data %>% filter(year.x == 2022) %>% ggplot(aes(x = lawtotal, y = rate)) +
  geom_point(size = 4, alpha = 0.7, color = "blue") +  
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed", color = "red") +  
  theme_minimal(base_size = 14) +  
  labs(title = "More Gun Laws have Lower Firearm Death Rates",
       x = "Total Gun Laws (LAWTOTAL)",
       y = "Firearm Deaths Rate")  



```

## Handgun Permit Laws Reduce Firearm Deaths?

-   This compares firearm deaths in states that require a handgun permit (PERMITH \~= 1) vs those that don't (PERMITH \~= 0).
-   This shows (PERMITH \~= 1) has a lower median death rate, it suggests that handgun permit laws reduce firearm deaths.

## Key Observations:

-   The blue boxplot (states requiring permits) has a lower median firearm death rate compared to the red boxplot (no permit required).
-   This suggests that states with handgun permit laws may have fewer firearm deaths on average.
-   Wider Spread of Death Rates in non-permit states (Red)

```{r}

# Boxplot
firearm_mortality_data %>% filter(year.x ==2022)%>% filter(year.x == 2022, !is.na(permith))%>%ggplot(aes(x = as_factor(permith), y = rate, fill = permith)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Boxplot with transparency
  geom_jitter(aes(color = permith), width = 0.2, alpha = 0.6, size = 2) +  # Add jittered points
  labs(title = "Impact of Handgun Permit Laws on Firearm Deaths",
       x = "Handgun Permit Required (0 = No, 1 = Yes)",
       y = "Firearm Deaths Rate") +
  theme_minimal() +
  theme(legend.position = "top", text = element_text(size = 14))

```
## Deaths Variability in Permit/Non-Permit States

-   No permit required states has a wider interquartile range and more extreme outliers indicating higher variability in firearm death rates.
-   The permit required states shows a more compact distribution with fewer extreme outliers suggesting handgun permit laws tend to lower firearm death rates.

## Strong Positive Correlation b/w Poverty and Firearm Deaths

-   The scatter plot below shows a strong positive correlation between poverty and firearm deaths.
-   States with higher poverty levels tend to experience more gun-related fatalities, suggesting that economic hardship contributes to gun violence.


```{r}
# Load required libraries
library(ggplot2)
library(dplyr)


firearm_mortality_data %>% filter(year.x ==2022)%>% filter(year.x == 2022, !is.na(state))

firearm_mortality_data <- firearm_mortality_data %>%
  left_join(census_df, by = c("state" = "name")) %>% mutate(firearm_mortality_data, per_capita_poverty = (poverty_count/population)*100)

#print(firearm_mortality_data)

# Create a scatter plot with trend line
firearm_mortality_data %>% ggplot(aes(x = per_capita_poverty, y = rate)) +
  geom_point(aes(color = per_capita_poverty), size = 3, alpha = 0.7) +  # Scatter plot points
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", size = 1.2) +  # Regression line
  scale_color_gradient(low = "blue", high = "red") +  # Color scale
  theme_minimal(base_size = 14) +
  labs(title = "Firearm Deaths vs Poverty",
       subtitle = "Strong Positive Correlation between Poverty and Firearm Deaths",
       x = "Per Capita Poverty",
       y = "Firearm Deaths Rate",
       color = "Poverty Level") +
  theme(
    panel.grid.major = element_line(color = "gray85"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )

```

## Do Permit Laws have Potential Protective Effect?

-   The boxplot visualizes the relationship between permit laws (PERMITLAW: 0 \~= No), 1 \~= Yes) and death rates.

```{r}

# Boxplot with jittered points
firearm_mortality_data %>% filter(year.x ==2022)%>% filter(year.x == 2022, !is.na(permitlaw))%>%ggplot(aes(x = as_factor(permitlaw), y = rate, fill = permitlaw))+
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +  # Boxplot with transparent fill
  geom_jitter(aes(color = permitlaw), width = 0.2, alpha = 0.6, size = 2) +  # Jittered points
  labs(
    title = "Impact of Permit Laws on Death Rates",
    subtitle = "Permit Laws and Gun-Related Death Rates – A Potential Protective Effect?",
    x = "Permit Law (0 = No, 1 = Yes)",
    y = "Death Rate"
  ) +
  #scale_fill_manual(values = c("#F8766D", "#00BFC4")) +  # Fix color mapping
  #scale_color_manual(values = c("#F8766D", "#00BFC4")) +  # Fix jitter colors
  theme_minimal() +
  theme(legend.position = "none")  # Remove redundant legend


```
##  Lower Median Death Rates in States with Permit Laws 

-   The central line inside each box represents the median (middle value).
-   If the median for PERMITLAW = 1 (states with permit laws) is lower than for PERMITLAW = 0, this suggests that permit laws might be associated with lower death rates.
-   Conversely, if they are similar, it suggests no strong association.

##  Wider Interquartile Range without Permit Laws States

-   The height of each box represents the middle 50% of data (IQR).
-   If PERMITLAW = 0 has a larger spread, it suggests more variation in death rates in states without permit laws.
-   A smaller spread for PERMITLAW = 1 could indicate that states with permit laws have more consistent (and possibly lower) death rates.

## More Outliers without Permit Laws States

-   The jittered points (dots) represent individual states.
-   If many high death rate outliers appear in the PERMITLAW = 0 group, it suggests that states without permit laws might experience more extreme cases of high gun-related deaths.

```{r}
mortality_breakdown_df
```


```{r}
# First join the datasets
firearm_mortality_data <- firearm_mortality_data %>%
  left_join(mortality_breakdown_df, by = c("code.x" = "code"))

# Then filter data for 2022
data_2022 <- firearm_mortality_data %>% 
  filter(year.x == 2022, !is.na(permitlaw))

# Create three separate plots with state dots
p1 <- ggplot(data_2022, aes(x = factor(permitlaw), y = rate, fill = factor(permitlaw))) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(text = state.x), width = 0.2, alpha = 0.7, size = 2) +
  labs(title = "Overall Rate") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(data_2022, aes(x = factor(permitlaw), y = suicide_rate, fill = factor(permitlaw))) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(text = state.x), width = 0.2, alpha = 0.7, size = 2) +
  labs(title = "Suicide Rate") +
  theme_minimal() +
  theme(legend.position = "none")

p3 <- ggplot(data_2022, aes(x = factor(permitlaw), y = homicide_rate, fill = factor(permitlaw))) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(text = state.x), width = 0.2, alpha = 0.7, size = 2) +
  labs(title = "Homicide Rate") +
  theme_minimal() +
  theme(legend.position = "none")

# Arrange plots in a 1x3 grid
library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 3, 
             top = "Firearm Mortality Rates by Permit Law Status (2022)")
```


