---
title: "Story3_Data608"
author: "Mubashira Qari"
date: 2025-2-23"
format: revealjs
editor: visual
---

```{r, warning=FALSE}


library(dplyr)
library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tools)
library(stringdist)
library(broom)
library(gridExtra)
library(gclus)
library(car)
library(VGAM)
library(MASS)
library(rpart.plot)
library(ggfortify)
library(gridExtra)
library(forecast)
library(fpp2)
library(fma)
library(kableExtra)
library(e1071)
library(mlbench)
library(ggcorrplot)
library(DataExplorer)
library(timeDate)
library(caret)
library(GGally)
library(corrplot)
library(RColorBrewer)
library(tibble)
library(tidyr)
library(reshape2)
library(mixtools)
library(skimr)
library(ggplot2)
library(dplyr)
library(usmap)   # For U.S. map plotting
library(sf)      # For spatial data handling 
library(viridis) # For color scales


# Loading Datasets
firearm_mortality_df <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/State_Overview-2019-2023.csv")


gun_deaths_by_demographics_df <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/Gun_Deaths_by_Demographics-2019-2023.csv")

# firearm_mortality_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/firearm_mortality2022.csv")

state_abbreviation_df <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_6/states_abbreviations.csv")
gun_deaths_by_demographics_df
```


```{r}

# Replacing value 'District of Columbia' by 'DC '
#firearm_mortality_df$STATE <- gsub("District of Columbia", "DC", firearm_mortality_df$STATE)


# Keeps all rows from firearm_mortality_data and fills in NA for missing matches in state_abbreviation
firearm_mortality_df <- firearm_mortality_df %>%
  left_join(state_abbreviation_df, by = c("State" = "State"))

gun_deaths_by_demographics_df <- gun_deaths_by_demographics_df %>%
  left_join(state_abbreviation_df, by = c("State" = "State"))


# Get working directory path
path <- getwd()

# Write dataframe to CSV

file_path <- "C:/Users/Uzma/Downloads/overview_df.csv"

write.csv(firearm_mortality_df, file = file_path, row.names = FALSE)

#firearm_mortality_df
gun_deaths_by_demographics_df


```


```{r}

#  Data Preparation

# Convert state names to uppercase for consistency
#overview_df$State <- toupper(overview_df$State)

# Remove "United States" row if present
overview_df <- overview_df %>% filter(State != "UNITED STATES")

# Check data structure
#str(overview_df)

```

## Choropleth Map: Firearm Mortality Rate Per 100,000 People:

-   This map highlights geographical disparities in firearm mortality across states.


```{r}

# Ensure the State column is correctly named and formatted
colnames(overview_df) <- tolower(colnames(overview_df))  # Convert column names to lowercase
overview_df <- overview_df %>% rename(state = state)  # Ensure the column is named "state"

# Convert state names to title case (e.g., "ALABAMA" → "Alabama")
overview_df$state <- tools::toTitleCase(tolower(overview_df$state))

library(plotly)

# Convert ggplot to interactive plot
ggplot_map <- plot_usmap(data = overview_df, values = "age.adjusted.rate..per.100.000.") +
  scale_fill_viridis_c(name = "Deaths per 100k") +
  labs(title = "Firearm Mortality Disparities Across the U.S. (2019-2023)",
       subtitle = "States with higher firearm deaths per 100,000 people",
       caption = "Data Source: CDC") +
  theme_minimal()

# Make the map interactive
ggplotly(ggplot_map)



```
## Key Findings:

-   The Southern and Western states tend to have higher firearm death rates.
-   Northeastern states (Massachusetts, New York, New Jersey) have the lowest mortality rates.

## Bar Chart: Comparing Firearm Deaths in Strict vs. Loose Gun Law States

-   Do Stricter Gun Laws Lead to Lower Firearm Mortality? A State Comparison


