---
title: "Assignment2_Data608"
author: "Mubashira Qari"
date: "2025-02-16"
output:
  slidy_presentation: default
  powerpoint_presentation: default
---

### Load Libraries

```{r}

library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tools)
library(stringdist)
library(broom)
library(gridExtra)
library(gclus)
library(car)
library(VGAM)
library(MASS)
library(rpart.plot)
library(ggfortify)
library(gridExtra)
library(forecast)
library(fpp2)
library(fma)
library(kableExtra)
library(e1071)
library(mlbench)
library(ggcorrplot)
library(DataExplorer)
library(timeDate)
library(caret)
library(GGally)
library(corrplot)
library(RColorBrewer)
library(tibble)
library(tidyr)
library(reshape2)
library(mixtools)
library(skimr)

```

### Loading Datasets

```{r}

unemployment_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_2/unemployment_rate.csv")

fed_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_2/fed_fund_rate.csv")

cpi_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_2/consumer_price_index.csv")

#print(unemployment_data)
#print(fed_data)
print(cpi_data)
```

### Remove the HALF1 and HALF2 columns

```{r}

cpi_df <- cpi_data[, !(names(cpi_data) %in% c("HALF1", "HALF2"))]

```

### Convert from wide format to long format using `reshape()`

```{r}
cpi_long <- reshape(cpi_df, 
                    varying = list(2:ncol(cpi_df)),  # All columns except "Year"
                    v.names = "CPI",  # New column name for CPI values
                    timevar = "Month",  # New column for Month names
                    times = names(cpi_df)[2:ncol(cpi_df)],  # Month names from column names
                    idvar = "Year",  # Keep Year column as identifier
                    direction = "long")
cpi_long
```
### Saving as CSV

```{r}
# Define the file path with filename and extension
#file_path <- "C:/Users/Uzma/Downloads/new_df.csv"
file_path <- "C:/Users/Uzma/CUNY-SPS-Assignments/Data_608/output_csv/processed_interest_rates.csv"
# Write dataframe to CSV
write.csv(cpi_long, file = file_path, row.names = FALSE)

# Confirm that the file was saved
print("File saved successfully!")
cpi_long

```

### Calculate percentage change and round to 2 decimal places

```{r}
cpi_df <- cpi_long %>%
  group_by(Month) %>%  # Group by month to ensure comparisons are within the same month
  mutate(`inflation_rate` = round(((CPI - lag(CPI)) / lag(CPI)) * 100, 2)) %>%
  ungroup()

cpi_df
```


### Create a new column 'Status' based on 'inflation_rate'

```{r}

cpi_df <- cpi_df %>%
  mutate(inflation_criteria = ifelse(inflation_rate > 2, "Not Achieved", "Achieved"))

cpi_df

```

### Remove rows where Year is 1999

```{r}

cpi_df <- cpi_df %>%
  filter(Year != 1999)  # Keep only rows where Year is NOT 1999

cpi_df

```

### Saving as CSV

```{r}
# Define the file path with filename and extension
#file_path <- "C:/Users/Uzma/Downloads/new_df.csv"
file_path <- "C:/Users/Uzma/CUNY-SPS-Assignments/Data_608/output_csv/inflation_rates.csv"
# Write dataframe to CSV
write.csv(cpi_df, file = file_path, row.names = FALSE)

# Confirm that the file was saved
print("File saved successfully!")
cpi_df

```

### Preparing FED Dataset

```{r}
fed_data
```



### Convert "observation_date" to a Date format

```{r}
fed_data <- fed_data %>%
  mutate(observation_date = mdy(observation_date))  # Converts MM/DD/YYYY format to Date type

fed_df <- fed_data
fed_df

```
### Extract Year and Month

```{r}
fed_df <- fed_df %>%
  mutate(Year = year(observation_date),
         Month = month(observation_date, label = TRUE, abbr = TRUE))  # Extract month name (Jan, Feb, etc.)

fed_df
```

### Saving as CSV

```{r}
# Define the file path with filename and extension
#file_path <- "C:/Users/Uzma/Downloads/new_df.csv"
file_path <- "C:/Users/Uzma/CUNY-SPS-Assignments/Data_608/output_csv/fed_rates.csv"
# Write dataframe to CSV
write.csv(fed_df, file = file_path, row.names = FALSE)

# Confirm that the file was saved
print("File saved successfully!")


```



### Creating Dataframe for Unemployment Rate dataset

```{r}
unemp_df <- unemployment_data

unemp_df
```
### Extract the Month from the Label column

```{r}
unemp_df <- unemp_df %>%
  mutate(Month = word(Label, 2))  # Extract the second word (month name) from "1999 Jan"

unemp_df
```
### Arrange dataset in chronological order

```{r}
unemp_df <- unemp_df %>%
  arrange(Year, Month, unemployment_rate)

unemp_df

```
### Create a new column 'Status' based on 'unemployment_rate'

```{r}

unemp_df <- unemp_df %>%
  mutate(unemp_criteria = ifelse(unemployment_rate > 5, "Not Achieved", "Achieved"))

unemp_df

```

### Saving as CSV

```{r}
# Define the file path with filename and extension
#file_path <- "C:/Users/Uzma/Downloads/new_df.csv"
file_path <- "C:/Users/Uzma/CUNY-SPS-Assignments/Data_608/output_csv/unemp_rates.csv"
# Write dataframe to CSV
write.csv(unemp_df, file = file_path, row.names = FALSE)

# Confirm that the file was saved
print("File saved successfully!")


```
### Visualizing FED's Mandate Fulfillment

The Federal Reserve (FED) has a dual mandate from Congress:

Stable Prices (Low Inflation) → Inflation rate around 2%

Maximum Employment (Low Unemployment) → Low unemployment (~4% or lower)

### Load Libraries

```{r}

library(ggplot2)  
library(dplyr)     
library(readr)     
library(scales)    

```

### Remove Year 1999

```{r}

unemp_df <- unemp_df %>% filter(Year != 1999)
cpi_df <- cpi_df %>% filter(Year != 1999)
fed_df <- fed_df %>% filter(Year != 1999)

```


### Data Preparation: Merge All Three Datasets

```{r}

merged_df <- unemp_df %>%
  inner_join(cpi_df, by = c("Year", "Month")) %>%
  inner_join(fed_df, by = c("Year", "Month"))

merged_df

```
### Visualization: Has the FED Fulfilled its Mandate?

### Unemployment, Inflation & Fed Funds Rate Trend

This line chart shows how the unemployment rate, inflation rate, and Fed Funds rate have changed over time.

```{r}

# Create a combined plot with dual y-axes
ggplot(merged_df, aes(x = as.Date(paste(Year, Month, "1", sep = "-"), "%Y-%b-%d"))) +
  geom_line(aes(y = unemployment_rate, color = "Unemployment Rate"), size = 1.2) +
  geom_line(aes(y = inflation_rate, color = "Inflation Rate"), size = 1.2) +
  geom_line(aes(y = fed_fund_rate, color = "Fed Funds Rate"), size = 1.2, linetype = "dashed") +
  scale_y_continuous(sec.axis = sec_axis(~., name = "Fed Funds Rate (%)")) +
  labs(title = "Unemployment & Inflation vs. Fed Funds Rate",
       x = "Year", 
       y = "Unemployment & Inflation (%)",
       color = "Legend") +
  theme_minimal()


```
### The Federal Reserve System has been given a dual mandate of pursuing the economic goals of maximum employment and price stability with a inflation rate of 2% over time and unemployment rate between 4% and 6% over time

### "Has the FED been able to fulfill the mandate given to it by Congress?"


### Unemployment Rate Over Time

```{r}
ggplot(merged_df, aes(x = as.Date(paste(Year, Month, "1", sep = "-"), "%Y-%b-%d"), y = unemployment_rate)) +
  geom_line(color = "blue", size = 1.2) +
  geom_hline(yintercept = 4, linetype = "dashed", color = "red") +
  labs(title = "Unemployment Rate Over Time",
       x = "Year", 
       y = "Unemployment Rate (%)") +
  theme_minimal()

```
Interpretation:

A dashed red line at 4% shows the employment mandate
If unemployment is below 4%, Fed is meeting its employment goal

### Inflation Rate Over Time

```{r}

ggplot(merged_df, aes(x = as.Date(paste(Year, Month, "1", sep = "-"), "%Y-%b-%d"), y = inflation_rate)) +
  geom_line(color = "green", size = 1.2) +
  geom_hline(yintercept = 2, linetype = "dashed", color = "red") +
  labs(title = "Inflation Rate Over Time",
       x = "Year", 
       y = "Inflation Rate (%)") +
  theme_minimal()

```
Interpretation:

A dashed red line at 2% shows the price stability mandate
If inflation stays near 2%, Fed is achieving price stability


To evaluate whether the Federal Reserve (FED) has met its unemployment mandate, we need to visualize unemployment trends over time and compare them with the "Achieved" vs. "Not Achieved" status.

```{r}
# Convert Month to numeric and create Date column
unemp_df <- unemp_df %>%
  mutate(
    Month = match(Month, month.abb),  # Convert "Jan" -> 1, "Feb" -> 2
    Date = as.Date(paste(Year, Month, "1", sep = "-"), format = "%Y-%m-%d"),
    unemployment_rate = as.numeric(unemployment_rate)  # Ensure numeric
  )


```


```{r}
ggplot(cpi_df, aes(x = factor(Year), fill = inflation_criteria)) +
  geom_bar(position = "stack") +  
  scale_fill_manual(values = c("Achieved" = "green", "Not Achieved" = "orange")) +
  scale_y_continuous(breaks = seq(0, 12, by = 1)) +  # Force whole numbers for months
  labs(title = "Inflation Mandate Achievement Over the Years",
       x = "Year",
       y = "Number of Months",
       fill = "Mandate Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```



```{r}
ggplot(unemp_df, aes(x = factor(Year), fill = unemp_criteria)) +
  geom_bar(position = "stack") +  
  scale_fill_manual(values = c("Achieved" = "blue", "Not Achieved" = "pink")) +
  scale_y_continuous(breaks = seq(0, 12, by = 1)) +  # Ensure y-axis is in whole numbers (0 to 12)
  labs(title = "Unemployment Mandate Achievement Over the Years",
       x = "Year",
       y = "Number of Months",
       fill = "Mandate Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

```

### Ensure merged_df is Properly Structured

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# Ensure the Month column is a character type
merged_df <- merged_df %>%
  mutate(Month = as.character(Month))

# Convert to long format for better visualization
long_df <- merged_df %>%
  pivot_longer(cols = c(inflation_criteria, unemp_criteria),
               names_to = "Mandate",
               values_to = "Status")

# Rename "Mandate" column for clarity
long_df <- long_df %>%
  mutate(Mandate = ifelse(Mandate == "inflation_criteria", "Inflation Mandate", "Unemployment Mandate"))



```

### Create the Stacked Bar Chart

### Now visualizing combined Inflation & Unemployment Mandate Achievements over the years.

```{r}

# Define custom colors for each category
colors <- c("Inflation Mandate.Achieved" = "green",
            "Inflation Mandate.Not Achieved" = "orange",
            "Unemployment Mandate.Achieved" = "green",
            "Unemployment Mandate.Not Achieved" = "blue")

# Create the stacked bar chart
ggplot(long_df, aes(x = factor(Year), fill = interaction(Mandate, Status))) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = colors) +
  scale_y_continuous(breaks = seq(0, 24, by = 2)) +  # Ensure whole number scale for months
  labs(title = "Mandate Achievement Over the Years (Inflation & Unemployment)",
       x = "Year",
       y = "Number of Months Achieved",
       fill = "Mandate Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability


```
### Explanation of the Visualization

Each Year is on the X-axis.

Y-axis represents months (0–12 per mandate, up to 24 total).

Stacked bars show whether mandates were Achieved or Not Achieved:

Green → Achieved (for both inflation & unemployment)
Blue → Unemployment Not Achieved
Orange → Inflation Not Achieved

### Final Outcome

This clearly illustrates how well the FED has met its dual mandate over time, highlighting patterns in inflation and unemployment stability.
