---
title: "Story2_Data608"
author: "Mubashira Qari"
date: 2025-2-23"
format: revealjs
editor: visual
---

```{r, warning=FALSE, echo=FALSE, include=FALSE}
#install.packages("tinytex")
#tinytex::install_tinytex()  # Install TinyTeX (if not installed)
#tinytex::tlmgr_install("koma-script")  # Install the missing KOMA-Script package


library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tools)
library(stringdist)
library(broom)
library(gridExtra)
library(gclus)
library(car)
library(VGAM)
library(MASS)
library(rpart.plot)
library(ggfortify)
library(gridExtra)
library(forecast)
library(fpp2)
library(fma)
library(kableExtra)
library(e1071)
library(mlbench)
library(ggcorrplot)
library(DataExplorer)
library(timeDate)
library(caret)
library(GGally)
library(corrplot)
library(RColorBrewer)
library(tibble)
library(tidyr)
library(reshape2)
library(mixtools)
library(skimr)


# Loading Datasets
unemployment_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_2/unemployment_rate.csv")

fed_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_2/fed_fund_rate.csv")

cpi_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week_2/consumer_price_index.csv")

#print(unemployment_data)
#print(fed_data)
#print(cpi_data)

# Remove the HALF1 and HALF2 columns
cpi_df <- cpi_data[, !(names(cpi_data) %in% c("HALF1", "HALF2"))]


# Convert from wide format to long format using `reshape()`
cpi_long <- reshape(cpi_df, 
                    varying = list(2:ncol(cpi_df)),  # All columns except "Year"
                    v.names = "CPI",  # New column name for CPI values
                    timevar = "Month",  # New column for Month names
                    times = names(cpi_df)[2:ncol(cpi_df)],  # Month names from column names
                    idvar = "Year",  # Keep Year column as identifier
                    direction = "long")


# Calculate percentage change and round to 2 decimal places
cpi_df <- cpi_long %>%
  group_by(Month) %>%  # Group by month to ensure comparisons are within the same month
  mutate(`inflation_rate` = round(((CPI - lag(CPI)) / lag(CPI)) * 100, 2)) %>%
  ungroup()

#cpi_df

# Create a new column 'Status' based on 'inflation_rate'
cpi_df <- cpi_df %>%
  mutate(inflation_criteria = ifelse(inflation_rate > 2, "Not Achieved", "Achieved"))

#cpi_df


# Remove rows where Year is 1999
cpi_df <- cpi_df %>%
  filter(Year != 1999)  # Keep only rows where Year is NOT 1999

#cpi_df


# Preparing FED Dataset
#fed_data

# Convert "observation_date" to a Date format
fed_data <- fed_data %>%
  mutate(observation_date = mdy(observation_date))  # Converts MM/DD/YYYY format to Date type

fed_df <- fed_data
#fed_df


# Extract Year and Month
fed_df <- fed_df %>%
  mutate(Year = year(observation_date),
         Month = month(observation_date, label = TRUE, abbr = TRUE))  # Extract month name (Jan, Feb, etc.)

#fed_df

# Creating Dataframe for Unemployment Rate dataset
unemp_df <- unemployment_data

#unemp_df

# Extract the Month from the Label column
unemp_df <- unemp_df %>%
  mutate(Month = word(Label, 2))  # Extract the second word (month name) from "1999 Jan"

#unemp_df

# Arrange dataset in chronological or
unemp_df <- unemp_df %>%
  arrange(Year, Month, unemployment_rate)

#unemp_df


# Create a new column 'Status' based on 'unemployment_rate'
unemp_df <- unemp_df %>%
  mutate(unemp_criteria = ifelse(unemployment_rate > 6, "Not Achieved", "Achieved"))

#unemp_df

```

## Visualizing FED's Mandate Fulfillment {.smaller}

-   The Federal Reserve (FED) has a dual mandate from Congress which according to economist is:
-   Stable Prices (Low Inflation) when Inflation rate is around 2% and maximum Employment (Low Unemployment) when low unemployment is below 5% - 6%

## Feature Engineering Criteria for inflation {.smaller}

-   If inflation_rate > (\~2%), it is classified as "Not Achieved" (inflation is too high).
-   Otherwise, it is classified as "Achieved" (inflation is at or below the target).
-   This new column inflation_criteria indicates whether the inflation target was met for that period.

```{r, warning=FALSE}
# Remove Year 1999
library(ggplot2)  
library(dplyr)     
library(readr)     
library(scales) 
unemp_df <- unemp_df %>% filter(Year != 1999)
cpi_df <- cpi_df %>% filter(Year != 1999)
fed_df <- fed_df %>% filter(Year != 1999)


# Data Preparation: Merge All Three Datasets
merged_df <- unemp_df %>%
  inner_join(cpi_df, by = c("Year", "Month")) %>%
  inner_join(fed_df, by = c("Year", "Month"))

#merged_df


# Creating a Status Field
merged_df <- merged_df %>%
  mutate(
    status = case_when(
      unemp_criteria == "Achieved" & inflation_criteria == "Achieved" ~ "Yes",
      unemp_criteria == "Not Achieved" & inflation_criteria == "Not Achieved" ~ "No",
      TRUE ~ "Partial Achieved"
    )
  )

#merged_df

```

## Feature Engineering Criteria for Unemployment {.smaller}

-   The unemp_criteria field is based on a threshold that determines whether the unemployment rate is at an acceptable level or too high. The criteria for assigning values to this field are as follows:
-   Achieved → If the unemployment rate is at or below (\~6%), it is considered within the acceptable range, meaning the economic goal for employment is met.
-   Not Achieved → If the unemployment rate is above (\~6%), it indicates high unemployment, meaning the target for employment has not been met.

## Has the FED fulfilled its mandate? {.smaller}

-   This line chart below shows how the unemployment rate, inflation rate, and Fed Funds rate have changed over time.


```{r}
# Load required libraries
library(ggplot2)   # For visualization
library(dplyr)     # For data manipulation
library(tidyr)     # For reshaping data

# Ensure the date column is correctly formatted
merged_df <- merged_df %>%
  mutate(date = as.Date(paste(Year, Month, "1", sep = "-"), "%Y-%b-%d"))

# Convert wide data into long format
merged_df_long <- merged_df %>%
  pivot_longer(cols = c(unemployment_rate, inflation_rate, fed_fund_rate),
               names_to = "Indicator",
               values_to = "Value")

# Define criteria lines data frame
criteria_lines <- data.frame(
  Indicator = c("inflation_rate", "unemployment_rate"),
  yintercept = c(2, 6)  # Inflation Rate = 2%, Unemployment Rate = 6%
)

# Create the faceted time series plot with criteria lines
ggplot(merged_df_long, aes(x = date, y = Value, color = Indicator)) +
  geom_line(size = 1.2) +
  
  # Add criteria lines only for Inflation and Unemployment plots
  geom_hline(data = criteria_lines, aes(yintercept = yintercept),
             color = c("red", "blue"), linetype = "dashed", size = 1) +

  labs(title = "Economic Indicators Over Time with Criteria Lines",
       x = "Year",
       y = "Value (%)",
       color = "Indicator") +
  theme_minimal() +
  facet_wrap( ~Indicator, scales = "free_y",nrow=3)  # Separate time series for each indicator

```

## Fed Funds Rate (Red Line) {.smaller}

-   The Fed Funds Rate represents the interest rate set by the Federal Reserve, influencing borrowing costs and monetary policy.

## Peaks (\~2000, 2006, 2019, 2023), Drops (\~2003, 2009, 2015, 2020) {.smaller}

-   Around 2000, rates were above 6%, likely due to the Federal Reserve's tightening to control inflation during the dot-com bubble.
-   Around 2006, rates again peaked (\~5.25%) before the 2008 financial crisis, as the Fed tried to curb housing market overheating.
-   Around 2019, rates rose (\~2.5%) as the Fed tightened policy post-Great Recession recovery.
-   2023 peak (\~5%) due to aggressive hikes to combat inflation after the COVID-19 stimulus spending.
-   2003: The Fed cut rates below 2% following the dot-com bubble burst and 9/11 economic slowdown.
-   2009: Sharp drop to nearly 0% as a response to the 2008 financial crisis.
-   2015: Rates remained near zero until gradual hikes began post-recession.
-   2020: Plunged back to near 0% due to the COVID-19 pandemic to stimulate the economy.

## Inflation Rate (Green Line) {.smaller}

-   The Inflation Rate measures the rise in prices over time, typically influenced by demand, supply shocks, and monetary policy.

## Peaks (\~2008, 2022), Drops (\~2009, 2015, 2020) {.smaller}

-   2008: Inflation rose above 5% due to high oil prices before collapsing during the financial crisis.
-   2022: Inflation spiked above 8%, driven by supply chain disruptions, post-COVID stimulus, and high consumer demand.
-   2009: Inflation briefly turned negative (\~-1%), indicating deflation after the financial crisis recession.
-   2015: Inflation stayed low (\~0%), reflecting slow economic recovery and low energy prices.
-   2020: Inflation dipped as the economy shut down due to COVID-19, before rebounding sharply in 2021-2022.

## Unemployment Rate (Blue Line) {.smaller}

-   The Unemployment Rate measures the percentage of people actively seeking jobs, reflecting labor market strength.

## Peaks (\~2010, 2020), Drops (\~2007, 2019, 2022) {.smaller}

-   2010: Unemployment spiked to \~10% after the 2008 financial crisis, as companies cut jobs during the Great Recession.
-   2020: Massive peak (\~15%) due to the COVID-19 pandemic lockdowns, which led to millions losing jobs almost overnight.
-   2007: Unemployment was low (\~4%) before the financial crisis.
-   2019: Unemployment fell to \~3.5%, the lowest in decades, signaling a strong labor market before COVID-19.
-   2022: After the pandemic shock, the labor market recovered rapidly, bringing unemployment back to \~3.5%.

## Unemployment Figure

```{r}
# Convert Month to numeric and create Date column
unemp_df <- unemp_df %>%
  mutate(
    Month = match(Month, month.abb),  # Convert "Jan" -> 1, "Feb" -> 2
    Date = as.Date(paste(Year, Month, "1", sep = "-"), format = "%Y-%m-%d"),
    unemployment_rate = as.numeric(unemployment_rate)  # Ensure numeric
  )

ggplot(cpi_df, aes(x = factor(Year), fill = inflation_criteria)) +
  geom_bar(position = "stack") +  
  scale_fill_manual(values = c("Achieved" = "green", "Not Achieved" = "orange")) +
  scale_y_continuous(breaks = seq(0, 12, by = 1)) +  # Force whole numbers for months
  labs(title = "Inflation Mandate Rarely Achieved Over the Years",
       x = "Year",
       y = "Number of Months",
       fill = "Mandate Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```

## Key Takeaways from "Inflation Mandate Rarely Achieved Over the Years" Chart {.smaller}

-   Inflation was rarely within target levels for sustained periods (mostly orange bars).
-   Post-crisis years (2009-2015) saw temporary improvements but lacked consistency.
-   Only 2016-2019 showed a relatively stable period of inflation control.
-   Recent years (2020-2024) have seen inflation mandate failures due to extreme economic shocks.
-   This highlights the Federal Reserve's ongoing struggle to maintain inflation within target levels, showing that inflation control is highly sensitive to external crises and policy shifts.

## Unemployment Mandate Achievement Over the Years {.smaller}

-   Each bar represents a full year (12 months), with colors indicating whether unemployment was within the target range for that month.
-   Blue (Achieved): Indicates months where unemployment was at or below the target.
-   Pink (Not Achieved): Represents months where unemployment exceeded the target, signaling economic distress.

## Unemployment Mandate Achievement Over the Years

```{r}
ggplot(unemp_df, aes(x = factor(Year), fill = unemp_criteria)) +
  geom_bar(position = "stack") +  
  scale_fill_manual(values = c("Achieved" = "blue", "Not Achieved" = "pink")) +
  scale_y_continuous(breaks = seq(0, 12, by = 1)) +  # Ensure y-axis is in whole numbers (0 to 12)
  labs(title = "Unemployment Mandate Achievement Over the Years",
       x = "Year",
       y = "Number of Months",
       fill = "Mandate Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

```

## Key Takeaways

-   Unemployment spikes during major economic downturns (2008-2014, 2020-2021) and takes years to recover.
-   Strong job markets (2000-2007, 2016-2019) show prolonged mandate achievement.
-   Crisis periods are clearly reflected in the extended pink bars, showing the challenge of labor market recovery.

## Conclusion:

-   The Federal Reserve’s employment mandate has historically been met during stable economic periods but fails significantly during recessions, requiring years of recovery.
