---
title: "Week_1_Assignment-Data_608"
author: "Mubashira Qari"
date: "2024-12-21"
output: html_document
---

### Load Libraries

```{r}

library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tools)
library(stringdist)

```


```{r}

IIJA_funding_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week%201/IIJA%20FUNDING%20AS%20OF%20MARCH%202023.csv")

state_population_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week%201/State_Population_Data.csv")


election_results_2020_data <- read.csv("https://raw.githubusercontent.com/uzmabb182/Data_608/refs/heads/main/Week%201/2020_Presidential_Election_Results.csv")

```


### View the First Few Rows of Data
```{r}

#print(IIJA_funding_data)

```

```{r}

# Check column names
colnames(IIJA_funding_data)

```

```{r}
# Renaming the columns

IIJA_funding_data <- IIJA_funding_data %>%
  rename(State_Name = State..Teritory.or.Tribal.Nation, Funding_Billions = Total..Billions.)

#print(IIJA_funding_data)
```


```{r}

# Convert column names to lowercase
colnames(IIJA_funding_data ) <- tolower(colnames(IIJA_funding_data ))

# View the modified data frame
#print(IIJA_funding_data )

```


```{r}
# Sort by population (ascending order)
IIJA_funding_data <- IIJA_funding_data[order(IIJA_funding_data$state_name),]

# View the sorted dataframe
#print(IIJA_funding_data)

```


```{r}

# Replace "Deleware" with "Delaware" conditionally
IIJA_funding_data$state_name[IIJA_funding_data$state_name == "Deleware"] <- "Delaware"

# View the updated DataFrame
#print(IIJA_funding_data)

```


```{r}

# Add index column using mutate
IIJA_funding_data <- IIJA_funding_data %>% mutate(index = row_number())

# View the updated DataFrame
#print(IIJA_funding_data)

```


```{r}

# Convert first letter of each word to uppercase using str_to_title
IIJA_funding_data$state_name <- str_to_title(IIJA_funding_data$state_name)

# View the updated DataFrame
#print(IIJA_funding_data)


```



```{r}

# Convert column names to lowercase
colnames(state_population_data ) <- tolower(colnames(state_population_data ))

# View the modified data frame
#print(state_population_data)

```


```{r}
# Sort by population (ascending order)
state_population_data <- state_population_data[order(state_population_data$name),]

# View the sorted dataframe
#print(state_population_data)

```



```{r}


# Convert first letter of each word to uppercase using str_to_title
state_population_data$name <- str_to_title(state_population_data$name)

# View the updated DataFrame
#print(state_population_data)

```



```{r}

# Convert column names to lowercase
colnames(election_results_2020_data ) <- tolower(colnames(election_results_2020_data ))

# View the modified data frame
#print(election_results_2020_data)

```

```{r}


# Convert first letter of each word to uppercase using str_to_title
election_results_2020_data$state <- str_to_title(election_results_2020_data$state)

# View the updated DataFrame
#print(election_results_2020_data)

```


```{r}

# Checking datatypes
#str(IIJA_funding_data)

#str(election_results_2020_data)

###str(state_population_data)

```


```{r}

# Convert 'state' column from integer to character
state_population_data$name <- as.character(state_population_data$name)

###str(state_population_data)

```



```{r}

# Merge on 'state_name' from df1 and 'state' from df2
result_df <- merge(IIJA_funding_data, state_population_data, by.x = "state_name", by.y = "name")

# View the merged dataframe
#print(result_df)

```


```{r}


# Merge on 'state_name' from df1 and 'state' from df2
result_df <- merge(IIJA_funding_data, state_population_data, by.x = "state_name", by.y = "name")

# View the merged dataframe
#print(result_df)

```



```{r}


# Merge on 'state_name' from df1 and 'state' from df2
merged_df <- merge(result_df, election_results_2020_data, by.x = "state_name", by.y = "state")

# View the merged dataframe
#print(merged_df)


```




```{r}

# Select specific columns
new_df <- merged_df %>% select(state_name, funding_billions, popestimate2020, trump_win, biden_win)

# View the new data frame
#print(new_df)

```


### Create Key Metrics for Analysis

### Funding Per state percentage

```{r}
# Add a new column
new_df <- new_df %>%
  mutate(funding_per_state_percentage = round((funding_billions / sum(funding_billions)) * 100, 2))

# Print the merged data frame
#print(new_df)

```


```{r}

# Add a new column
new_df <- new_df %>%
  mutate(population_per_state_percentage = round((popestimate2020 / sum(popestimate2020)) * 100, 2))

# Print the merged data frame
#print(new_df)


```



```{r}

# Add a new columns for testing criteria
new_df <- new_df %>%
  mutate(population_per_state_plus5percent = population_per_state_percentage + (population_per_state_percentage * 0.05))


new_df <- new_df %>%
  mutate(population_per_state_minus5percent = population_per_state_percentage - (population_per_state_percentage * 0.05))

new_df <- new_df %>%
  mutate(Equitable = ifelse(funding_per_state_percentage > population_per_state_minus5percent & 
                            funding_per_state_percentage < population_per_state_plus5percent, 
                            "Yes", "No"))
new_df <- new_df %>%
  mutate(bias = ifelse(biden_win == 1 & Equitable == "No", "Yes", "No"))

# Print the merged data frame
#print(new_df)

```

### Summary of Why These Fields Are Created

The following new fields are created to analyze whether funding allocation is equitable and to detect potential political bias in how resources are distributed.

### population_per_state_plus5percent & population_per_state_minus5percent

### Purpose:

These fields define an acceptable funding range (+/- 5%) around each state's population percentage.

### Why?

If funding is within ±5% of the population share, it is considered fair (equitable).

If funding falls outside this range, the state may be overfunded or underfunded.

population_per_state_plus5percent → Upper threshold (+5% of population share).

population_per_state_minus5percent → Lower threshold (-5% of population share).

### Equitable (Is the funding fair?)

### Purpose:

This field determines whether a state's funding is proportional to its population based on the ±5% rule.

### Why?

If funding falls within the ±5% range, the state is labeled "Yes" (Equitable).

If funding falls outside the range, it is labeled "No" (Inequitable).

This ensures we have an objective way to measure fairness in funding allocation.

### bias (Is there political bias in funding?)

### Purpose:

This field identifies political bias in funding allocation, specifically if Democratic (Biden-won) states received inequitable funding.

Why?

A state is considered biased ("Yes") if:

The state voted for Biden (biden_win == 1)

The state did NOT receive equitable funding (Equitable == "No")

Otherwise, the state is labeled as "No" (No bias detected).

This helps detect whether Democratic states were treated differently compared to Republican states.


### Is the Allocation Equitable Based on Population?

### Funding Allocation vs. Population

This chart compares funding per state percentage to population per state percentage. If funding is fair, bars should align with the population.

```{r}
ggplot(new_df, aes(x = reorder(state_name, population_per_state_percentage), 
                   y = funding_per_state_percentage, 
                   fill = Equitable)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("Yes" = "blue", "No" = "red")) +
  labs(title = "Funding Allocation vs. Population Percentage",
       x = "State",
       y = "Funding Percentage",
       fill = "Equitable") +
  theme_minimal()

```


Interpretation:

Blue bars ("Yes") → States where funding aligns with population.

Red bars ("No") → States overfunded or underfunded, indicating bias.

If many states are red, the allocation is not equitable.


### Funding vs. Population Percentage

If funding is fair, points should align in a linear trend.

```{r}

ggplot(new_df, aes(x = population_per_state_percentage, 
                   y = funding_per_state_percentage, 
                   color = Equitable)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  scale_color_manual(values = c("Yes" = "blue", "No" = "red")) +
  labs(title = "Funding Allocation vs. Population Percentage",
       x = "Population Percentage",
       y = "Funding Percentage",
       color = "Equitable") +
  theme_minimal()


```

Interpretation:

A strong trend line suggests fair allocation.

Scattered points with many "No" (Red) indicate funding was not proportional.

This visualizes bias more clearly than the bar chart.

### Does Allocation Favor the Biden Administration?

### Funding vs. Biden Vote Share

If Democratic states receive more funding, the trend line will be upward-sloping.

```{r}

ggplot(new_df, aes(x = as.numeric(biden_win), 
                   y = funding_per_state_percentage, 
                   color = bias)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) +
  labs(title = "Bias in Funding Allocation vs. Biden Vote",
       x = "Biden Win (1 = Won, 0 = Lost)",
       y = "Funding Percentage",
       color = "Bias") +
  theme_minimal()


```

Interpretation:

Upward slope → Biden-won states received more funding.

No slope → Funding was politically neutral.

If most red points (bias) are above, funding favored Biden states.

### Funding by Political Lean

If Biden-won states received higher funding, the blue box will be taller.


```{r}

ggplot(new_df, aes(x = factor(biden_win), 
                   y = funding_per_state_percentage, 
                   fill = factor(biden_win))) +
  geom_boxplot() +
  scale_fill_manual(values = c("0" = "red", "1" = "blue")) +
  labs(title = "Funding Allocation by Political Lean",
       x = "Biden Win (1 = Won, 0 = Lost)",
       y = "Funding Percentage",
       fill = "Biden Win") +
  theme_minimal()

```

Interpretation:

If the blue box (Biden states) is higher, funding favored Democrats.

If both boxes are similar, funding was neutral.

Outliers in red box may indicate some Republican states were underfunded.

Conclusion

1- Was allocation equitable?
The bar chart & scatter plot show many states in red ("No"), proving funding was not proportional to population.
Some states received far more or far less than expected, suggesting bias or inefficiencies.

2️- Was there political bias?
The scatter plot & boxplot show that Democratic states received more funding than Republican states.
This suggests possible favoritism toward the Biden administration.

 Final Verdict: Allocation was not equitable and appears to favor Democratic-leaning states.
 
 
 ### Number of States with Bias
 
```{r}
ggplot(new_df, aes(x = bias, fill = bias)) +
  geom_bar() +
  scale_fill_manual(values = c("Yes" = "red", "No" = "blue")) +
  labs(title = "Bias in Funding Allocation",
       x = "Bias Present",
       y = "Number of States",
       fill = "Bias") +
  theme_minimal()

```
 Interpretation:

More red ("Yes") bars → Bias is present in multiple Democratic states.

More blue ("No") bars → No strong pattern of bias.

Summary
Now bias is only "Yes" when Biden-won states were not equitably funded.

Republican states (biden_win == 0) do not count, even if they were treated inequitably
.
This correctly isolates political favoritism toward Democratic states.

```{r}

# Define the file path with filename and extension
file_path <- "C:/Users/Uzma/Downloads/new_df.csv"

# Write dataframe to CSV
write.csv(new_df, file = file_path, row.names = FALSE)

# Confirm that the file was saved
print("File saved successfully!")

```

```{r}
# Select specific columns
group_df <- new_df %>% select(state_name, funding_billions, popestimate2020, trump_win, biden_win)

# View the new data frame
###print(group_df)

```



```{r}

# Calculate total funding and population
total_funding <- sum(group_df$funding_billions, na.rm = TRUE)
total_population <- sum(group_df$popestimate2020, na.rm = TRUE)

# Create a new table with grouped states and percentages
grouped_table <- group_df %>%
  group_by(trump_win, biden_win) %>%
  summarize(
    trump_funding_percentage = round(sum(ifelse(trump_win == 1, funding_billions, 0), na.rm = TRUE) / total_funding * 100, 2),
    biden_funding_percentage = round(sum(ifelse(biden_win == 1, funding_billions, 0), na.rm = TRUE) / total_funding * 100, 2),
    trump_population_percentage = round(sum(ifelse(trump_win == 1, popestimate2020, 0), na.rm = TRUE) / total_population * 100, 2),
    biden_population_percentage = round(sum(ifelse(biden_win == 1, popestimate2020, 0), na.rm = TRUE) / total_population * 100, 2)
  ) %>%
  ungroup()

# Print the new grouped table with rounded percentages
print(grouped_table)

```


```{r}

# Define the file path with filename and extension
file_path <- "C:/Users/Uzma/Downloads/grouped_table.csv"

# Write dataframe to CSV
write.csv(grouped_table, file = file_path, row.names = FALSE)

# Confirm that the file was saved
print("File saved successfully!")

```

### Comparing Funding vs. Population Distribution

After calculating the funding percentage and population percentage for states that voted for Trump (Republican) or Biden (Democrat), I'll create insightful visualizations to highlight potential biases.

### Bar Chart

```{r}
# Convert data to long format for easy visualization
grouped_long <- grouped_table %>%
  pivot_longer(cols = c(trump_funding_percentage, biden_funding_percentage, 
                        trump_population_percentage, biden_population_percentage), 
               names_to = "Category", 
               values_to = "Percentage")

# Create labels for clarity
grouped_long$Group <- ifelse(grepl("trump", grouped_long$Category), "Trump-Won States", "Biden-Won States")
grouped_long$Metric <- ifelse(grepl("funding", grouped_long$Category), "Funding Allocation", "Population Percentage")

# Create the bar chart
ggplot(grouped_long, aes(x = Group, y = Percentage, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Funding Allocation" = "red", "Population Percentage" = "blue")) +
  labs(title = "Funding Allocation vs. Population Share by Political Affiliation",
       x = "Political Affiliation (2020 Election)",
       y = "Percentage of Total",
       fill = "Category") +
  theme_minimal()


```

Analysis of the Bar Chart

If funding allocation closely matches population share, then the distribution is likely fair.

If Biden states receive significantly more funding than their population share, it suggests possible bias in allocation.

If Trump states receive less funding despite a larger population share, it may indicate underfunding relative to need.


### Share of Total Funding by Political Affiliation

This pie chart visualizes which party’s states received more funding.

### Pie Chart

```{r}
# Create a summarized data frame for pie chart
funding_pie <- grouped_table %>%
  summarize(
    Trump = sum(trump_funding_percentage),
    Biden = sum(biden_funding_percentage)
  ) %>%
  pivot_longer(cols = everything(), names_to = "Party", values_to = "Funding_Share")

# Pie Chart
ggplot(funding_pie, aes(x = "", y = Funding_Share, fill = Party)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +  # Convert to a pie chart
  scale_fill_manual(values = c("Trump" = "red", "Biden" = "blue")) +
  labs(title = "Share of Total Funding by Political Affiliation",
       fill = "Party") +
  theme_minimal()


```


Analysis of the Pie Chart

If Biden states dominate the pie chart, this suggests funding may have favored Democratic states.

If the chart is balanced, then allocation was politically neutral.

### Funding vs. Population for Biden vs. Trump States

This scatter plot shows whether Biden states received disproportionately more funding.

### Scatter Plot

```{r}

ggplot(grouped_table, aes(x = trump_population_percentage, y = trump_funding_percentage, color = "Trump States")) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", linetype = "dashed", color = "red") +
  geom_point(aes(x = biden_population_percentage, y = biden_funding_percentage, color = "Biden States"), size = 4) +
  geom_smooth(aes(x = biden_population_percentage, y = biden_funding_percentage), 
              method = "lm", linetype = "dashed", color = "blue") +
  scale_color_manual(values = c("Trump States" = "red", "Biden States" = "blue")) +
  labs(title = "Funding Allocation vs. Population Share for Trump and Biden States",
       x = "Population Share (%)",
       y = "Funding Share (%)",
       color = "Political Group") +
  theme_minimal()

```
Analysis of the Scatter Plot

A stronger funding trend for Biden states (blue trendline sloping higher) suggests favoritism.

If Trump states show a weaker funding trend, they may have been underfunded relative to their population size.

### Final Analysis

### Key Findings:

### Funding was not equitably distributed based on population

Biden states received a higher percentage of funding than their population share, while Trump states received less funding relative to their population.

### Funding appears to favor Biden-won states

The bar chart & pie chart show that Biden states got a higher proportion of total funding.

The scatter plot confirms a stronger funding trend for Democratic states, suggesting possible political bias.


### Conclusion:

This analysis provides strong evidence that funding allocation was NOT purely based on population and may have favored Biden-won states.









